Single Bivalent Level Analysis

### 2020-01-22 doc/Bivalent_Analysis.Rmd
first pass at analysis of Mixed models, outlined other predictions, made some notes in other files

---
title: "Single Bivalent Level Analysis"
author: "April Peterson"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}

library(knitr)
library(ggplot2)
#library(pwr)
library(plyr)
library(lattice)
library(dplyr)
library(raster)
library(lme4)
library(nlme)
library(RLRsim)
library(reshape2)

setwd("C:/Users/alpeterson7/Documents/MLH1repo/")
#setwd("C:/Users/April/Desktop/MLH1repo")
#load(file="data/MLH1/MLH1_data_setup.RData")

#load Clean BivData
load(file="data/CleanBivData_1.21.20.RData")

#load other data files, MLH1
load(file="C:/Users/alpeterson7/Documents/MLH1repo/data/MLH1/MLH1_data_setup_1.5.20.RData")#added batch18

@@

@@

@@

@@

Interference, measured in the physical units, has been shown to lack sexual dimorphism (petkov).

@@

@@

$$mouse \ average \ SC  \ length ~=~ subsp * sex + rand(strain* sex) + \varepsilon $$

Above is the model we are testing evolution of SC lengths.

```{r MM.SC.Length, echo=FALSE}

#use the merged DF to filter by age 
AP_mouse_table_w.Ages$strain <- AP_mouse_table_w.Ages$strain.x
AP_mouse_table_w.Ages$sex <- AP_mouse_table_w.Ages$sex.x
AP_mouse_table_w.Ages$subsp <- AP_mouse_table_w.Ages$subsp.x
#main file, make everything unordered factors
AP_mouse_table_w.Ages <- add_species(AP_mouse_table_w.Ages)

DF.HetC.MixedModel <- AP_mouse_table_w.Ages[AP_mouse_table_w.Ages$species == "M.musculus",]

#check all the strains and male ages, should be the same for MM

DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel[DF.HetC.MixedModel$Ncells > 9,]
DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel.HQ[DF.HetC.MixedModel.HQ$age.weeks < 25,]
DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel.HQ[!(is.na(DF.HetC.MixedModel.HQ$mouse) | DF.HetC.MixedModel.HQ$mouse==""), ]

#mouse.level.table
#make this in datasetup
mouse.avs <- ddply(Curated_BivData, c("mouse"), summarise,
                        Nmice = length(unique(mouse)),
                        Ncells  = length(Obj.ID),
                        mean_SC = as.numeric(format(round(  mean(chromosomeLength), 3 ), nsmall=3) ),
                        cV = cv(chromosomeLength),
                        var = format(round(   var(chromosomeLength),3), nsmall=3),
                        sd   = round(sd(chromosomeLength), 3),
                        se   = round(sd / sqrt(Ncells), 3)
)

mouse.avs <- add_sex(mouse.avs)
mouse.avs <- add_strain(mouse.avs)
mouse.avs <- add_subsp(mouse.avs)

#unorder the factors, but figure our
mouse.avs$strain <- unclass(mouse.avs$strain)
mouse.avs$strain <- as.factor(mouse.avs$strain)

mouse.avs$subsp <- unclass(mouse.avs$subsp)
mouse.avs$subsp <- as.factor(mouse.avs$subsp)

mouse.avs$sex <- unclass(mouse.avs$sex)
mouse.avs$sex <- as.factor(mouse.avs$sex)

BivData.MixedModel.HQ <- lme(mean_SC ~ subsp * sex, data=mouse.avs, random=list(strain=pdDiag(~sex) ) )

as.data.frame( round(  summary(BivData.MixedModel.HQ)$tTable, 6 ) )

#sex is significant! with prediction

@@

The results seem to indicate that sex is a significant factor. Consider writing a subsampling approach (randomize / permute a data set of BivData)

@@

@@

DO I have anything to say....

@@

```{r show.IFD, echo=FALSE, warning=FALSE}

#first draft of the IFD comparisons

#for now remove the 
Curated_BivData.22 <- Curated_BivData[Curated_BivData$hand.foci.count == 2,]
Curated_BivData.22 <- Curated_BivData.22[(!is.na(Curated_BivData.22$hand.foci.count)),]

first.IFD.plot_ABS <- ggplot(Curated_BivData.22, aes(y = IFD1_ABS, x= strain, color=mouse))+geom_jitter()+ggtitle("Interfocal Distances, ABS")+facet_wrap(sex~subsp, scales = "free")+ylim(c(20,160))+theme(legend.position="none")
first.IFD.plot_ABS

@@

#+theme(legend.position="none")
first.IFD.plot_PER

@@

mixed model analysis for IFD (interference)

$$mouse \ average \ IFD ~=~ subsp * sex + rand(strain* sex) + \varepsilon $$

$$mouse \ average \ normalized IFD ~=~ subsp * sex + rand(strain* sex) + \varepsilon $$

I tested 2 versions of the mixed model for this flavor of trait, raw IFD and normalized IFD measure

@@

#set up data set for the mixed model
#Curated_BivData

#use the merged DF to filter by age 
AP_mouse_table_w.Ages$strain <- AP_mouse_table_w.Ages$strain.x
AP_mouse_table_w.Ages$sex <- AP_mouse_table_w.Ages$sex.x
AP_mouse_table_w.Ages$subsp <- AP_mouse_table_w.Ages$subsp.x
#main file, make everything unordered factors
AP_mouse_table_w.Ages <- add_species(AP_mouse_table_w.Ages)

DF.HetC.MixedModel <- AP_mouse_table_w.Ages[AP_mouse_table_w.Ages$species == "M.musculus",]

#check all the strains and male ages, should be the same for MM

DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel[DF.HetC.MixedModel$Ncells > 9,]
DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel.HQ[DF.HetC.MixedModel.HQ$age.weeks < 25,]
DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel.HQ[!(is.na(DF.HetC.MixedModel.HQ$mouse) | DF.HetC.MixedModel.HQ$mouse==""), ]

#mouse.level.table
#make this in datasetup
mouse.avs.table1 <- ddply(Curated_BivData, c("mouse"), summarise,
                        Nmice = length(unique(mouse)),
                        Ncells  = length(Obj.ID),
                        mean_SC = as.numeric(format(round(  mean(chromosomeLength, na.rm = TRUE), 3 ), nsmall=3) ),

                   mean_IFD_ABS = as.numeric(format(round(  mean(IFD1_ABS, na.rm = TRUE), 3 ), nsmall=3) ),

>>>>

---
title: "Single Bivalent Level Analysis"
author: "April Peterson"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
  pdf_document: default
---

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
#library(pwr)
library(plyr)
library(lattice)
library(dplyr)
library(raster)
library(lme4)
library(nlme)
library(RLRsim)
library(reshape2)

#setwd("C:/Users/alpeterson7/Documents/MLH1repo/")
#setwd("C:/Users/April/Desktop/MLH1repo")
#load(file="data/MLH1/MLH1_data_setup.RData")

setwd("~./MLH1repo/")
#load Clean BivData
load(file="data/CleanBivData_1.21.20.RData")

#load other data files, MLH1
load(file="~./MLH1repo/data/MLH1_data_setup_1.5.20.RData")#added batch18

#load(file="C:/Users/alpeterson7/Documents/MLH1repo/data/MLH1/MLH1_data_setup_1.5.20.RData")#added batch18

@@

# Chromosome Size effect

In an attempt to deal with the chromosome effect size that is know to effect total rate and CO position, we try to make a dataset of the longest 25% SC from a cell.

The majority of cells from the BivData set do not have the complete completement of bivalents, (ie they are missing up to 5) -- which makes it hard to use a absolute length threshold since some cells have shorter or longer bivalents compared to the while. So an absolute measure might catch the lower 25% of some cells.

I'd like to figure out how to best seperate out the true 4th quartille bivalents (longest 25%), even with this missing data. double check that 4th quartille is top 5/4

Could I use the total SC? (I would be warry: can't ensure that they apply the same skeletonize algorithm/parameters and the XY is also included in that count and will mess everything up.)

outline logic for filtering / choosing BivData set of almost whole cells

- 16 to 20 bivalents in the cells

```{r chrm.size.effect, echo=FALSE}

#calq the mouse means
manual_summary_stats <- ddply(whole.cell.data_manual, c("fileName"), summarise,

                        mean_SC = as.numeric(format(round(  mean(SC.length, na.rm = TRUE), 3 ), nsmall=3) ),
                        median_SC = median(SC.length, na.rm = TRUE),
                        Qar.4 = quantile(SC.length, na.rm = TRUE)[[4]]
#                        var_SC = format(round(   as.numeric(var(SC.length)),3), nsmall=3)

)

manual_summary_stats <- add_mouse(manual_summary_stats)
manual_summary_stats <- add_strain(manual_summary_stats)
manual_summary_stats <- add_sex(manual_summary_stats)
manual_summary_stats <- add_subsp(manual_summary_stats)

###
#calq the mouse means
Biv_Data_mean_table <- ddply(almost_whole_cell_BivData, c("fileName"), summarise,

                        mean_SC = as.numeric(format(round(  mean(chromosomeLength), 3 ), nsmall=3) ),
                        median_SC = median(chromosomeLength),
                        Qar.4 = quantile(chromosomeLength)[[4]],

                        var_SC = format(round(   as.numeric(var(chromosomeLength)),3), nsmall=3)
        #add Q4

)
Biv_Data_mean_table <- add_mouse(Biv_Data_mean_table)
Biv_Data_mean_table <- add_strain(Biv_Data_mean_table)
Biv_Data_mean_table <- add_sex(Biv_Data_mean_table)
Biv_Data_mean_table <- add_subsp(Biv_Data_mean_table)

show.the.cells_females_dom <- ggplot(whole.cell.data_manual[(whole.cell.data_manual$sex == "female") & (whole.cell.data_manual$subsp == "Dom"),], aes(x=fileName, y= SC.length, color=strain))+geom_point()+

   geom_point(data=manual_summary_stats[(manual_summary_stats$sex == "female") &(manual_summary_stats$subsp == "Dom"),],aes(y=mean_SC, x=fileName, col="red", size=2))+
  geom_point(data=manual_summary_stats[(manual_summary_stats$sex == "female") & (manual_summary_stats$subsp == "Dom"),],aes(y=median_SC, x=fileName, col="black", size=1.5))+

   geom_point(data=manual_summary_stats[(manual_summary_stats$sex == "female") & (manual_summary_stats$subsp == "Dom"),],aes(y=Qar.4, x=fileName, col="black", size=2.6), shape=4)+

  facet_wrap(~strain, scales="free")+ylim(c(30,200))+ggtitle("Female whole cell manual measures")
show.the.cells_females_dom
#PWD seems to have the biggest variance across cells

almost_cells_DOM_f <- ggplot(almost_whole_cell_BivData[(almost_whole_cell_BivData$sex == "female") & (almost_whole_cell_BivData$subsp == "Dom"),], aes(x=fileName, y= chromosomeLength, color=strain))+
  geom_point()+ 
  geom_point(data=Biv_Data_mean_table[(Biv_Data_mean_table$sex == "female") & (Biv_Data_mean_table$subsp == "Dom"),],aes(y=mean_SC, x=fileName, col="red", size=2))+
  geom_point(data=Biv_Data_mean_table[(Biv_Data_mean_table$sex == "female") & (Biv_Data_mean_table$subsp == "Dom"),],aes(y=median_SC, x=fileName, col="black", size=1.5))+
   geom_point(data=Biv_Data_mean_table[(Biv_Data_mean_table$sex == "female") & (Biv_Data_mean_table$subsp == "Dom"),],aes(y=Qar.4, x=fileName, col="black", size=2.6), shape=4)+

  facet_wrap(~strain, scales="free")+ylim(c(30,200))+ggtitle("Dom female, almost whole cell \nMean, median, 4th Quartille ploted")
almost_cells_DOM_f

#### Musc female

show.the.cells_females_musc <- ggplot(whole.cell.data_manual[(whole.cell.data_manual$sex == "female") & (whole.cell.data_manual$subsp == "Musc"),], aes(x=fileName, y= SC.length, color=strain))+geom_point()+

   geom_point(data=manual_summary_stats[(manual_summary_stats$sex == "female") &(manual_summary_stats$subsp == "Musc"),],aes(y=mean_SC, x=fileName, col="red", size=2))+
  geom_point(data=manual_summary_stats[(manual_summary_stats$sex == "female") & (manual_summary_stats$subsp == "Musc"),],aes(y=median_SC, x=fileName, col="black", size=1.5))+

   geom_point(data=manual_summary_stats[(manual_summary_stats$sex == "female") & (manual_summary_stats$subsp == "Musc"),],aes(y=Qar.4, x=fileName, col="black", size=2.6), shape=4)+

  facet_wrap(~strain, scales="free")+ylim(c(30,200))+ggtitle("Female Musc whole cell manual measures")
show.the.cells_females_musc

almost_cells_Musc_f <- ggplot(almost_whole_cell_BivData[(almost_whole_cell_BivData$sex == "female") & (almost_whole_cell_BivData$subsp == "Musc"),], aes(x=fileName, y= chromosomeLength, color=strain))+

  geom_point()+ 
  geom_point(data=Biv_Data_mean_table[(Biv_Data_mean_table$sex == "female") & (Biv_Data_mean_table$subsp == "Musc"),],aes(y=mean_SC, x=fileName, col="red", size=2))+
  geom_point(data=Biv_Data_mean_table[(Biv_Data_mean_table$sex == "female") & (Biv_Data_mean_table$subsp == "Musc"),],aes(y=median_SC, x=fileName, col="black", size=1.5))+
   geom_point(data=Biv_Data_mean_table[(Biv_Data_mean_table$sex == "female") & (Biv_Data_mean_table$subsp == "Musc"),],aes(y=Qar.4, x=fileName, col="black", size=2.6), shape=4)+

  facet_wrap(~strain, scales="free")+ylim(c(30,200))+ggtitle("Musc female, almost whole cell \nMean, median, 4th Quartille ploted")
almost_cells_Musc_f

### Dom Male
show.the.cells_male_dom <- ggplot(whole.cell.data_manual[(whole.cell.data_manual$sex == "male") & (whole.cell.data_manual$subsp == "Dom"),], aes(x=fileName, y= SC.length, color=strain))+geom_point()+

   geom_point(data=manual_summary_stats[(manual_summary_stats$sex == "male") &(manual_summary_stats$subsp == "Dom"),],aes(y=mean_SC, x=fileName, col="red", size=2))+
  geom_point(data=manual_summary_stats[(manual_summary_stats$sex == "male") & (manual_summary_stats$subsp == "Dom"),],aes(y=median_SC, x=fileName, col="black", size=1.5))+

   geom_point(data=manual_summary_stats[(manual_summary_stats$sex == "male") & (manual_summary_stats$subsp == "Dom"),],aes(y=Qar.4, x=fileName, col="black", size=2.6), shape=4)+

  facet_wrap(~strain, scales="free")+ylim(c(30,150))+ggtitle("Male Dom whole cell manual measures")
show.the.cells_male_dom

almost_cells_Dom_m <- ggplot(almost_whole_cell_BivData[(almost_whole_cell_BivData$sex == "male") & (almost_whole_cell_BivData$subsp == "Dom"),], aes(x=fileName, y= chromosomeLength, color=strain))+

  geom_point()+ 
  geom_point(data=Biv_Data_mean_table[(Biv_Data_mean_table$sex == "male") & (Biv_Data_mean_table$subsp == "Dom"),],aes(y=mean_SC, x=fileName, col="red", size=2))+
  geom_point(data=Biv_Data_mean_table[(Biv_Data_mean_table$sex == "male") & (Biv_Data_mean_table$subsp == "Dom"),],aes(y=median_SC, x=fileName, col="black", size=1.5))+
   geom_point(data=Biv_Data_mean_table[(Biv_Data_mean_table$sex == "male") & (Biv_Data_mean_table$subsp == "Dom"),],aes(y=Qar.4, x=fileName, col="black", size=2.6), shape=4)+

  facet_wrap(~strain, scales="free")+ylim(c(30,150))+ggtitle("Dom male, almost whole cell \nMean, median, 4th Quartille ploted")
almost_cells_Dom_m

#MUSC male
show.the.cells_male_musc <- ggplot(whole.cell.data_manual[(whole.cell.data_manual$sex == "male") & (whole.cell.data_manual$subsp == "Musc"),], aes(x=fileName, y= SC.length, color=strain))+geom_point()+

   geom_point(data=manual_summary_stats[(manual_summary_stats$sex == "male") &(manual_summary_stats$subsp == "Musc"),],aes(y=mean_SC, x=fileName, col="red", size=2))+
  geom_point(data=manual_summary_stats[(manual_summary_stats$sex == "male") & (manual_summary_stats$subsp == "Musc"),],aes(y=median_SC, x=fileName, col="black", size=1.5))+

   geom_point(data=manual_summary_stats[(manual_summary_stats$sex == "male") & (manual_summary_stats$subsp == "Musc"),],aes(y=Qar.4, x=fileName, col="black", size=2.6), shape=4)+

  facet_wrap(~strain, scales="free")+ylim(c(30,150))+ggtitle("Musc Dom whole cell manual measures")
show.the.cells_male_musc

almost_cells_Musc_m <- ggplot(almost_whole_cell_BivData[(almost_whole_cell_BivData$sex == "male") & (almost_whole_cell_BivData$subsp == "Musc"),], aes(x=fileName, y= chromosomeLength, color=strain))+

  geom_point()+ 
  geom_point(data=Biv_Data_mean_table[(Biv_Data_mean_table$sex == "male") & (Biv_Data_mean_table$subsp == "Musc"),],aes(y=mean_SC, x=fileName, col="red", size=2))+
  geom_point(data=Biv_Data_mean_table[(Biv_Data_mean_table$sex == "male") & (Biv_Data_mean_table$subsp == "Musc"),],aes(y=median_SC, x=fileName, col="black", size=1.5))+
   geom_point(data=Biv_Data_mean_table[(Biv_Data_mean_table$sex == "male") & (Biv_Data_mean_table$subsp == "Musc"),],aes(y=Qar.4, x=fileName, col="black", size=2.6), shape=4)+

  facet_wrap(~strain, scales="free")+ylim(c(30,150))+ggtitle("Musc male, almost whole cell \nMean, median, 4th Quartille ploted")
almost_cells_Musc_m

#the min is pretty stable

#show.the.cells_male <- ggplot(whole.cell.data_manual[whole.cell.data_manual$sex == "male",], aes(x=fileName, y= SC.length, color=strain))+geom_point()+

#   geom_point(data=manual_summary_stats[manual_summary_stats$sex == "male",],aes(y=mean_SC, x=fileName, col="red", size=2))+

#  geom_point(data=manual_summary_stats[manual_summary_stats$sex == "male",],aes(y=median_SC, x=fileName, col="black", size=1.5))+

#   geom_point(data=manual_summary_stats[manual_summary_stats$sex == "male",],aes(y=Qar.4, x=fileName, col="black", size=2.6), shape=4)+

 # facet_wrap(~strain, scales="free")+ylim(c(30,150))+ggtitle("Male whole cell manual measures")
#show.the.cells_male

#almost_whole_cell_BivData
#almost_whole_cell_wsb <- almost_whole_cell_BivData[almost_whole_cell_BivData$strain == "WSB",]
#almost_whole_cell_DOM <- almost_whole_cell_BivData[almost_whole_cell_BivData$subsp == "Dom",]
#almost_whole_cell_MUSC <- almost_whole_cell_BivData[almost_whole_cell_BivData$subsp == "Musc",]

All the plots above show the distrtributions of manually whole cell measured SC lengths compared to the SC length distributions from the automated bivalent data. It shows the amount of within cell variance across strains. There is a bit of variance across the SC length distributions in the PWD females.

I'm purposing to take all the automated biv measures above the cell's 4th quartille as a dataset for 'longestest' bivalents which we would predict to exclude the shortest chromsomes which would have stronger chromsome size effects.

But this dataset might be noisy, given the amount of variance in the SC length distributions across cells (PWD females, WSB females). But it's easier to just run another set of analysis.

```{r long.chrm.dataset, echo=FALSE, eval=FALSE}

#how many biv's are from cells with >15 bivalents

al.15_bivs <- cell.level.BivData_table[cell.level.BivData_table$Nbiv >15, ]#150 cells have al 15

long_BivData_set <- Curated_BivData[ Curated_BivData$fileName %in% al.15_bivs$fileName, ] #across those 150 cells are 2654 biv measures

#table(potential.data.set$category)
long_BivData_set$fileName <- as.character(long_BivData_set$fileName)
Biv_Data_mean_table$fileName <- as.character(Biv_Data_mean_table$fileName)

#pull out the biv's which exceed the cell's Q4
#for each line of long_BivData_set
# check if it's SC length is greater than the Q4 for the given cell

long_BivData_set99 <- long_BivData_set

for(  i in 1:length( long_BivData_set99$fileName)  ){
  long_BivData_set99$cell.Q4[i] <- Biv_Data_mean_table[ which(Biv_Data_mean_table$fileName == long_BivData_set99$fileName[i]), ]
}

for(  k in 1:length( long_BivData_set99$fileName)  ){
long_BivData_set99$above.Q4[k] <- ifelse( (long_BivData_set99$cell.Q4 < long_BivData_set99$chromosomeLength), "yes", "no")
}
#This isn't working either

#for(i in 1:length( long_BivData_set$fileName)  ){ 
  # print(i)
#  current_obj_fileName <- long_BivData_set$fileName[i]
#  current_obj_SC_length <- long_BivData_set$chromosomeLength[i]  
#  current.Q4 <- Biv_Data_mean_table$Qar.4[which(Biv_Data_mean_table$fileName == current_obj_fileName) ]
# current.Q4 <- Biv_Data_mean_table$Qar.4[which(Biv_Data_mean_table$fileName == long_BivData_set$fileName[i]) ]
#  long_BivData_set$above.Q4 <-ifelse( (current.Q4 < current_obj_SC_length), 1, 0)
#  print( paste( c(current.Q4, current_obj_fileName, current_obj_SC_length), sep = "_" ) )
#}
#this loop isn't working

The dataset is unbalanced in some ways -- integrating manual measures might fix. MSM under rep, since most cell had high overlap.

@@

I will use two different mixed models using lme() and lmer(), because the different models have limitations for reporting p values ect.

@@

```{r clean.biv.data4.MM, echo=FALSE}

#remove all the non-used strains!!
Curated_BivData <- Curated_BivData[Curated_BivData$strain != "AST",]
Curated_BivData <- Curated_BivData[Curated_BivData$strain != "TOM",]
Curated_BivData <- Curated_BivData[Curated_BivData$strain != "CZECH",]
Curated_BivData$strain <- droplevels(Curated_BivData$strain)
#MOLF

#non-ordered factors, 
Curated_BivData$subsp <- factor(Curated_BivData$subsp, ordered = FALSE)
Curated_BivData$sex <- factor(Curated_BivData$sex, ordered = FALSE)
Curated_BivData$strain <- factor(Curated_BivData$strain, ordered = FALSE)

#all numbers, numbers
#all relavent factors unordered

#the mouse averages will be used

mouse.avs_4MM <- ddply(Curated_BivData, c("mouse"), summarise,
                        Nmice = length(unique(mouse)),
                        Ncells  = length(Obj.ID),
                        mean_SC = as.numeric(format(round(  mean(chromosomeLength), 3 ), nsmall=3) ),
                        cV = cv(chromosomeLength),
                        var = as.numeric(format(round(   var(chromosomeLength),3), nsmall=3) ),
                        sd   = round(sd(chromosomeLength), 3),
                        se   = round(sd / sqrt(Ncells), 3),
                       #IFD things
                #  mean_IFD_ABS = as.numeric(format(round(  mean(IFD1_ABS, na.rm = TRUE), 3 ), nsmall=3) ),
                # mean_IFD_PER = as.numeric(format(round(  mean(IFD1_PER, na.rm = TRUE), 3 ), nsmall=3) ),

                 mean_IFD.2CO_ABS = mean(IFD1_ABS[which(hand.foci.count == 2)], na.rm = TRUE),
                 mean_IFD.2CO_PER = mean(IFD1_PER[which(hand.foci.count == 2)], na.rm = TRUE),

                       #CO position things, requires the melt?
              norm.first.foci.pos =  mean(PER_Foci_1[which(hand.foci.count == 1)])

)

#vari
mouse.avs_4MM <- add_sex(mouse.avs_4MM)
mouse.avs_4MM <- add_strain(mouse.avs_4MM)
mouse.avs_4MM <- add_subsp(mouse.avs_4MM)
#unordered
mouse.avs_4MM$subsp <- factor(mouse.avs_4MM$subsp, ordered = FALSE)
mouse.avs_4MM$sex <- factor(mouse.avs_4MM$sex, ordered = FALSE)
mouse.avs_4MM$strain <- factor(mouse.avs_4MM$strain, ordered = FALSE)

#MAKE the same version of the TABLE but for the longest_biv_Dataset (when I get around to making it)

#code if males are high or low?

In the chunk above the mouse averages table is made -- may need to add all the extra metrics (IFD, .

@@

3) Interference, measured in the physical units, has been shown to lack sexual dimorphism (petkov).

@@

## Predictions, male polymorphism

What predictions do I have for how traits will differ between the low and high Musc mouse strains?

A) **SC lengths** will be longer for high Rec

B.1) **Interfernce/IFD** will be shorter in high Rec

or

B.2) **Interference/IFD** will be longer in high Rec males, sis-co-ten metric will be maximized. (given prelim results from Beth's data)

C) It's unknown if **1CO normalized positions** between high and low musc strains, will differ. (no good predictions come to mind)

Using the same MM framework we would expect the random strain effect would be significant for these predictions.

Use just the male (Musc male data)

$$mouse \ average \ SC  \ length ~=~ rand(strain) + \varepsilon $$

$$mouse \ average \ IFD \ trait  ~=~ rand(strain) + \varepsilon $$
The IFD traits are:

- IFD_ABS

- IFD_PER

$$mouse \ average \ CO \ position \ trait  ~=~ rand(strain) + \varepsilon $$

The CO position traits are:

- 1CO position

- centromere and telomere distance

- sis-co-ten

@@

$$mouse \ average \ SC  \ length ~=~ subsp * sex + rand(strain*sex) + \varepsilon $$

To estimate random effects, I'm using lmer (not lme where I could do the rand(strain*sex)). So now it's just random(strain).

Above is the model we are testing evolution of SC lengths.  remember for the MM use the table **mouse.avs_4MM**

```{r MM_2.SC.Length, echo=FALSE}
#get a list of the old mice to double check
#old.mice <- DF.HetC.MixedModel.HQ$mouse[DF.HetC.MixedModel.HQ$age.weeks < 25,]
#MIXED MODEL
lme_MM_SC.length <- lme(mean_SC ~ subsp * sex, data= mouse.avs_4MM, random=list(strain=pdDiag(~sex) ) )
as.data.frame( round(  summary(lme_MM_SC.length)$tTable, 6 ) )
#p

@@

I used the mixed model with the lme()

$$mouse \ average \ SC  \ length ~=~ subsp * sex + rand(strain) + \varepsilon $$

```{r MM_2_SC.length, echo=FALSE}

#as.data.frame( round(  summary(BivData.MixedModel.HQ)$tTable, 6 ) )

#can't make the 
SC_length.MM <- lmer(mean_SC ~ subsp * sex + (1|strain), data=mouse.avs_4MM, REML = TRUE)

anova(SC_length.MM)

thing1  <- summary(SC_length.MM)
thing2 <- drop1(SC_length.MM, test="Chisq")

rand_SC.Length_MM <- exactRLRT(SC_length.MM)
rand_SC.Length_MM

#this pvalue for ran is 0.0027 -- the strain effect 
```

@@

### Heterochiasmy Prediction

> Is sex a significant effect for SC length? (as predicted)

Sex is a significant effect for SC length
The results seem to indicate that sex is a significant factor. Consider writing a subsampling approach (randomize / permute a data set of BivData)

> Is the random Effect of strain an effect on SC length?

The exactRLRT() test indicates that the random strain effect might be significant, p= `r rand_SC.Length_MM$p.value`.

### Male polymorphism Predcition

Preidction, High rec males have longer SC.

```{r male.poly.test.MM, echo=FALSE, eval=FALSE}
#t.test comparisons of PWD_SC vs (KAZ, SKIVE, )
#MSM_SC vs MOLF_SC

#2. logistic regression for sc length and chrm class?
#if the prediction / significance

#isolate data
mouse.avs_4MM_male <- mouse.avs_4MM[mouse.avs_4MM$sex == "male",]
mouse.avs_4MM_M.male <- mouse.avs_4MM_male[mouse.avs_4MM_male$subsp == "Musc",]

mouse.avs_4MM_M.male$mouse <- factor(mouse.avs_4MM_M.male$mouse, ordered = FALSE)

#build and test MM
lme_MM_SC_length_male.poly <- lme(mean_SC ~ mouse + 1|strain, data= mouse.avs_4MM_M.male)#this syntax not working
as.data.frame( round(  summary(lme_MM_SC.length)$tTable, 6 ) )

lmer_MM_SC_length_male.poly <- lmer(mean_SC ~ (1|strain), data=mouse.avs_4MM_M.male, REML = TRUE)

anova(lmer_MM_SC_length_male.poly)

rand_SC.Length_MM.male.poly <- exactRLRT(lmer_MM_SC_length_male.poly)#this won't work either...hmm

I can't get the above mixed models to work for testing high vs low rec groups

```{r logreg, echo=FALSE}

mouse.avs_4MM_male <- mouse.avs_4MM[mouse.avs_4MM$sex == "male",]
mouse.avs_4MM_M.male <- mouse.avs_4MM_male[mouse.avs_4MM_male$subsp == "Musc",]

#divide musc males into groups
mouse.avs_4MM_male$Rec.group <- ifelse(grepl("PWD", mouse.avs_4MM_male$strain), 1, 
                                    ifelse(grepl("MSM", mouse.avs_4MM_male$strain), 1, 0))

log.reg_SC.length <- glm(Rec.group ~ mean_SC, 
                            data=mouse.avs_4MM_male, family=binomial(link="logit"))

log.reg_SC.length_Musc <- glm(Rec.group ~ mean_SC, 
                            data=mouse.avs_4MM_male[mouse.avs_4MM_male$subsp == "Musc",], family=binomial(link="logit"))

#the summary for both these models is the same ... should I be worried?

#mean SC slightly sig-- these are all mean SC--should I try distinguishing by chrm classes?
```

When all male mice are used, the predictive power is greater, than when just the Musc strains are used. When, just the Musc strain are used, The mouse mean SC is slightly significant in predicting if a mouse is in the high or low (should I consider running on female too?)

> Is the prediction, high rec musc male strains have long SC met?

In a logistic regression, mouse average SC length is slightly predictive telling if a mouse is in a high or low Rec strain.  I counldn't get the Mixed models working for the male polymorphism predictions...

@@

@@

```{r show.IFD, echo=FALSE, warning=FALSE}
colors_of_strains <- c('WSB'= "#56B4E9",'G' ='cadetblue','LEW'= 'lightblue','PERC'= 'blue',
                       #blues

                       'PWD'= 'red2',
                       'SKIVE'= 'coral1','KAZ'= 'coral4', 'TOM'= 'indianred1','AST' = 'red4',
                       'CZECH'= 'indianred4',

                       'MSM'= 'deeppink', 'MOLF'= 'hotpink', #mol are pinks, bc they are fancy mice

                    'CAST'='green','HMI'= 'forestgreen', #greens

                    'SPRET'='gold','SPIC'= 'goldenrod1', #browns? color of mounds?
                    'CAROLI'='gold3',

                    "F1"='grey', "other" = 'grey'
                    )
#first draft of the IFD comparisons
#make the mouse boxplots -- instead of geom_points

#Only use 2CO bivalents
Curated_BivData.22 <- Curated_BivData[Curated_BivData$hand.foci.count == 2,]
Curated_BivData.22 <- Curated_BivData.22[(!is.na(Curated_BivData.22$hand.foci.count)),]

#boxplots
boxplots_first.IFD.plot_ABS <- ggplot(Curated_BivData.22, aes(y = IFD1_ABS, x= mouse, color=strain))+geom_boxplot(alpha=.2)+
  geom_jitter(alpha=.2)+
  ggtitle("Interfocal Distances, ABS")+
  scale_color_manual(values=colors_of_strains)+  facet_wrap(sex~subsp, scales = "free")+ylim(c(20,160))+theme(legend.position="none")
boxplots_first.IFD.plot_ABS
#order is out of wack, but -- good for now

boxplots_first.IFD.plot_PER <- ggplot(Curated_BivData.22, aes(y = IFD1_PER, x= mouse, color=strain))+geom_boxplot(alpha=.2)+
  geom_jitter(alpha=.2)+
  ggtitle("Interfocal Distances, ABS")+
  scale_color_manual(values=colors_of_strains)+  facet_wrap(sex~subsp, scales = "free")+theme(legend.position="none")+ylim(c(0,1))
boxplots_first.IFD.plot_PER

first.IFD.plot_ABS <- ggplot(Curated_BivData.22, aes(y = IFD1_ABS, x= strain, color=mouse))+geom_jitter()+ggtitle("Interfocal Distances, ABS")+facet_wrap(sex~subsp, scales = "free")+ylim(c(20,160))+theme(legend.position="none")

@@

#+theme(legend.position="none")

@@

mixed model analysis for IFD (interference), the first set of models are made with the lme() functions.

$$mouse \ average \ IFD ~=~ subsp * sex + rand(strain*sex) + \varepsilon $$

```{r MM_1.SC.Length, echo=FALSE}
#get a list of the old mice to double check
#old.mice <- DF.HetC.MixedModel.HQ$mouse[DF.HetC.MixedModel.HQ$age.weeks < 25,]
#MIXED MODEL

#remove the SKIVE male line
mouse.avs_4MM <- mouse.avs_4MM[!(is.na(mouse.avs_4MM$mean_IFD.2CO_ABS) | mouse.avs_4MM$mean_IFD.2CO_ABS==""), ]

lme_MM_IFD_ABS <- lme(mean_IFD.2CO_ABS ~ subsp * sex, data= mouse.avs_4MM, random=list(strain=pdDiag(~sex) ) )
as.data.frame( round(  summary(lme_MM_IFD_ABS)$tTable, 6 ) )
#p

lme_MM_IFD_PER <- lme(mean_IFD.2CO_PER ~ subsp * sex, data= mouse.avs_4MM, random=list(strain=pdDiag(~sex) ) )
as.data.frame( round(  summary(lme_MM_IFD_PER)$tTable, 6 ) )
```

### second set of Mixed Models

These seond set of models are made using the lmer()

$$mouse \ average \ IFD ~=~ subsp * sex + rand(strain) + \varepsilon $$

$$mouse \ average \ normalized IFD ~=~ subsp * sex + rand(strain) + \varepsilon $$

I tested 2 versions of the mixed model for this flavor of trait, raw IFD and normalized IFD measure. The tables below are from anova( for the lmer model )

@@

IFD_ABS.MM <- lmer(mean_IFD.2CO_ABS ~ subsp * sex + (1|strain), data=mouse.avs_4MM, REML = TRUE)

anova(IFD_ABS.MM)

#thing1  <- summary(IFD_ABS.MM)
IFD.ABS_chi <- drop1(IFD_ABS.MM, test="Chisq") 
IFD.ABS_chi

rand_test_IFD.ABS.MM <- exactRLRT(IFD_ABS.MM)
rand_test_IFD.ABS.MM

IFD_PER.MM <- lmer(mean_IFD.2CO_PER ~ subsp * sex + (1|strain), data=mouse.avs_4MM, REML = TRUE)

anova(IFD_PER.MM)

Chis.sp_IFD.PER.MM <- drop1(IFD_PER.MM, test="Chisq") 
Chis.sp_IFD.PER.MM

rand_test_IFD.PER.MM <- exactRLRT(IFD_PER.MM)
rand_test_IFD.PER.MM
